A White Noise Approach to Evolutionary Ecology

### 2020-04-07 .A White Noise Approach To Evolutionary Ecology/draft.Rmd
expanded volperts argument

where $\bar x(t)$ and $\sigma^2(t)$ are respectively the mean trait and phenotypic variance at time $t\geq0$. Second, we assume selection is determined by the growth rate $m(h,x)$ that is differentiable with respect to both arguments and is bounded above by some $r\in\mathbb{R}$ for all $x\in\mathbb{R}$ and $h\geq0$. We denote by $K$ an operator that takes functions such as $\nu(x,t)$ as its argument and returns other functions that may still depend on the spatial variable $x$ and the temporal variable $t$. For our application we will be focused on operators $K$ that can be written as $K(\nu)(x,t)=\int_{-\infty}^{+\infty}\kappa(x,y)\nu(y,t)dy$ for some non-negative and bounded bivariate function $\kappa$. We set $h=K(\nu)(x,t)$ to account for nonlocal effects of the abundance density $\nu(x,t)$ on the fitness of individuals with trait value $x$ [@Volpert2014]. An example of a nonlocal effect is the competition for a common resource among all individuals regardless of their trait value so that individuals with differing trait values effect each others fitness. The term nonlocal comes from interpreting the trait value as spatial location. From here on we abbreviate $m(h,x)=m(K(\nu)(x,t),x)$ to just $m(\nu,x)$. Third, we assume mutation is captured by diffusion with coefficient $\frac{\mu}{2}$. Now that these technicalities have been treated we can model the demographic dynamics of a population and the dynamics of a quantitative character simultaneously by the PDE

@@

with the initial condition $\nu(x,0)$ described above. This PDE is semilinear due to the dependency of the growth rate $m(\nu,x)$ on the solution $\nu(x,t)$ and is referred to as a scalar reaction-diffusion equation [@lawrenceevans2010]. Interpreting the trait value $x$ as spatial location and ignoring nonlocal effects, which can be satisfied by setting $K(\nu)=\nu$, this model of spatially distributed population dynamics has been intensely studied [for a review, see @Cantrell2004]. When $\mu=0$, equation (\ref{eq1}) can be seen as an analog of equation (\ref{first}) for a continuum of types. By assuming mutation acts as diffusion, the effect of mutation causes $\nu(x,t)$ to flatten out over time. In fact, if the growth rate is constant across $x$, then this model of mutation will cause $\nu(x,t)$ to converge to a flat line as $t\to\infty$. Although clearly an idealized representation of biological reality, this model is sufficiently general to capture a large class of dynamics including density dependent growth and frequency dependent selection. As an example, logistic growth combined with quadratic stabilizing selection can be captured using the growth rate 

@@

Recall that $m(\nu,x)$ is shorthand for $m(K(\nu)(x,t),x)$. That is, $m:[0,+\infty)\times\mathbb{R}\to\mathbb{R}$. Following our assumptions of the main text, we have that $m(h,x)$ is differentiable with respect to both $x$ and $h$ and there exists $r\in\mathbb{R}$ such that $m(h,x)\leq r$ across all $h\geq0$ and $x\in\mathbb{R}$. As in the main text, we also assume the initial condition $\nu(x,0)$ satisfies
\begin{equation}
N(0)=\int_{\mathbb{R}}\nu(x,0)dx<+\infty,
\end{equation}
\begin{equation}
-\infty<\bar x(0)=\int_{\mathbb{R}}xp(x,0)dx<+\infty,
\end{equation}
\begin{equation}
\sigma^2(0)=\int_{\mathbb{R}}(x-\bar x(0))^2p(x,0)dx<+\infty,
\end{equation}
where $p(x,0)=\nu(x,0)/N(0)$, and we consider the PDE
\begin{equation}\label{PDE_SM}
\dot\nu(x,t)=m(\nu,x)\nu(x,t)+\frac{\mu}{2}\Delta\nu(x,t).
\end{equation}

@@

Here we show that solutions to the PDE described in the main text are continuous in time and space, twice continuously differentiable with respect to space and differentiable with respect to time. Our approach is an adaptation of that taken by Volpert [-@Volpert2014].

Recall that the PDE we consider is of the form
\begin{equation}\label{PDE_supp}
\dot\nu(x,t)=m(\nu,x)\nu(x,t)+\frac{\mu}{2}\Delta\nu(x,t),
\end{equation}
where $m(\nu,x)$ bounded above by $r\in\mathbb{R}$ is shorthand for $m(K(\nu)(x,t),x)$ and $K(\nu)(x,t)=\int_\mathbb{R}\kappa(x,y)\nu(y,t)dy$ for some non-negative, bounded $\kappa$. Denoting $L^1(\mathbb{R})$ the Lebesgue space of integrable functions on $\mathbb{R}$ (ie, the set of functions $f:\mathbb{R}\to\mathbb{R}$ that satisfy $\|f\|_1=\int_\mathbb{R}|f(x)|dx<+\infty$ under the equivalence relation $f=g$ iff $\|f-g\|_1=0$), we assume $\nu(\cdot,0)\in L^1(\mathbb{R})$. Under these assumptions, we have already proved in SM \S\ref{finite} that $\nu(\cdot,t)\in L^1(\mathbb{R})$ for all $t\geq0$. Before proving the regularity of $\nu(x,t)$, we review the approach taken by Volpert [-@Volpert2014].

@@

In chapter 9 \S2.2 of Volpert [-@Volpert2014] complimentary conditions are assumed where $\kappa(x,y)=\varphi(x-y)$ for some integrable function $\varphi$ and initial condition $\nu(\cdot,0)\in L^\infty(\mathbb{R})$, where $L^\infty(\mathbb{R})$ is the Lebesgue space of bounded functions on $\mathbb{R}$ (ie, the set of functions $f:\mathbb{R}\to\mathbb{R}$ satisfying $\|f\|_\infty=\sup_{x\in\mathbb{R}}|f(x)|<+\infty$ under the equivalence relation $f=g$ iff $\|f-g\|_\infty=0$). Instead of the growth term $m(\nu,x)\nu(x,t)$ of PDE (\ref{PDE_supp}), Volpert worked with $f(\nu(x,t),K(\nu)(x,t))$ for some function $f:\mathbb{R}^2\to\mathbb{R}$ differentiable with respect to both arguments and satisfying, for some $q\geq0$,
\begin{equation}
f(x,y)\leq q(1+|x|+|y|), \ \forall \ x,y\in\mathbb{R}.
\end{equation}
Under these conditions, Volpert used semigroup theory to prove the solution $\nu(x,t)$ takes values in $L^\infty(\mathbb{R})$ for all $t\geq0$ and is differentiable with respect to time and twice continuously differentiable with respect to space. In particular, defining $\nu(t)$ as the function $x\to\nu(x,t)$ and keeping $\Phi(x,t)$ as defined in SM \S\ref{finite}, Volpert states the semigroup $\{S(t)\}_{t\geq0}$ given by 
\begin{equation}
S(t)\nu(0)=\int_0^t\frac{\mu}{2}\Delta\nu(x,s)ds=\int_\mathbb{R}\Phi(x-y,t)\nu(y,0)dy, \ \ \forall \ \ t\geq0, \ \nu(0)\in L^\infty(\mathbb{R})
\end{equation}
satisfies $\|S(t)\nu(0)\|_\infty\leq\|\nu(0)\|_\infty$ for all $t\geq0$ and $\nu(0)\in L^\infty(\mathbb{R})$. Then, by setting
\begin{equation}
F(v)=f\left(v(x),\int_\mathbb{R}\varphi(x-y)v(y)dy\right), \ \ \forall \ \ v\in L^\infty(\mathbb{R}),
\end{equation}
Volpert notes that $F$ is locally Lipschitz continuous on $L^\infty(\mathbb{R})$ and in particular $\|F(v)\|_\infty\leq q\|v\|_\infty$. Hence, under Volperts conditions, solutions to PDE (\ref{PDE_supp}) can be written as
\begin{equation}\label{volp_sol}
\nu(t)=S(t)\nu(0)+\int_0^tS(t-s)F(\nu(s))ds\in L^\infty(\mathbb{R}) \ \ \forall \ \ t\geq0, \ \nu(0)\in L^\infty(\mathbb{R}).
\end{equation}
Thanks to properties of the semigroup $\{S(t)\}_{t\geq0}$, solutions $\nu(t)$ given by (\ref{volp_sol}) satisfy the claimed regularity conditions.

@@

To adapt Volperts approach to our situation, note that $\|S(t)\nu(0)\|_1=\|\nu(0)\|_1$ for all $\nu(0)\in L^1(\mathbb{R})$ and, setting
\begin{equation}
F(v)=m\left(\int_\mathbb{R}\kappa(x,y)v(y)dy,x\right)v(x), \ \ \forall \ \ v\in L^1(\mathbb{R}),
\end{equation}
we have $\|F(v)\|_1\leq r\|v\|_1$ for all $v\in L^1(\mathbb{R})$. Hence, using the same arguments employed by Volpert [-@Volpert2014], we have that solutions to PDE (\ref{PDE_supp}) under our assumptions can be written as
\begin{equation}
\nu(t)=S(t)\nu(0)+\int_0^tS(t-s)F(\nu(s))ds\in L^1(\mathbb{R}) \ \ \forall \ \ t\geq0, \ \nu(0)\in L^1(\mathbb{R})
\end{equation}
which therefore satisfy the claimed regularity conditions.

>>>>

where $\bar x(t)$ and $\sigma^2(t)$ are respectively the mean trait and phenotypic variance at time $t\geq0$. Second, we assume selection is determined by the growth rate $m(h,x)$ that is differentiable with respect to both arguments and is bounded above by some $r\in\mathbb{R}$ for all $x\in\mathbb{R}$ and $h\geq0$. We denote by $K$ an operator that takes functions such as $\nu(x,t)$ as its argument and returns other functions that may still depend on the spatial variable $x$ and the temporal variable $t$. For our application we will be focused on operators $K$ that can be written as $(K\nu)(x,t)=\int_{-\infty}^{+\infty}\kappa(x,y)\nu(y,t)dy$ for some non-negative and bounded bivariate function $\kappa$. We set $h=(K\nu)(x,t)$ to account for nonlocal effects of the abundance density $\nu(x,t)$ on the fitness of individuals with trait value $x$ [@Volpert2014]. An example of a nonlocal effect is the competition for a common resource among all individuals regardless of their trait value so that individuals with differing trait values effect each others fitness. The term nonlocal comes from interpreting the trait value as spatial location. From here on we abbreviate $m(h,x)=m\big((K\nu)(x,t),x\big)$ to just $m(\nu,x)$. Third, we assume mutation is captured by diffusion with coefficient $\frac{\mu}{2}$. Now that these technicalities have been treated we can model the demographic dynamics of a population and the dynamics of a quantitative character simultaneously by the PDE

@@

with the initial condition $\nu(x,0)$ described above. This PDE is semilinear due to the dependency of the growth rate $m(\nu,x)$ on the solution $\nu(x,t)$ and is referred to as a scalar reaction-diffusion equation [@lawrenceevans2010]. Interpreting the trait value $x$ as spatial location and ignoring nonlocal effects, which can be satisfied by setting $K\nu=\nu$, this model of spatially distributed population dynamics has been intensely studied [for a review, see @Cantrell2004]. When $\mu=0$, equation (\ref{eq1}) can be seen as an analog of equation (\ref{first}) for a continuum of types. By assuming mutation acts as diffusion, the effect of mutation causes $\nu(x,t)$ to flatten out over time. In fact, if the growth rate is constant across $x$, then this model of mutation will cause $\nu(x,t)$ to converge to a flat line as $t\to\infty$. Although clearly an idealized representation of biological reality, this model is sufficiently general to capture a large class of dynamics including density dependent growth and frequency dependent selection. As an example, logistic growth combined with quadratic stabilizing selection can be captured using the growth rate 

@@

Recall that $m(\nu,x)$ is shorthand for $m\big((K\nu)(x,t),x\big)$. That is, $m:[0,+\infty)\times\mathbb{R}\to\mathbb{R}$. Following our assumptions of the main text, we have that $m(h,x)$ is differentiable with respect to both $x$ and $h$ and there exists $r\in\mathbb{R}$ such that $m(h,x)\leq r$ across all $h\geq0$ and $x\in\mathbb{R}$. As in the main text, we also assume the initial condition $\nu(x,0)$ satisfies
\begin{equation}
N(0)=\int_{\mathbb{R}}\nu(x,0)dx<+\infty,
\end{equation}
\begin{equation}
-\infty<\bar x(0)=\int_{\mathbb{R}}xp(x,0)dx<+\infty,
\end{equation}
\begin{equation}
\sigma^2(0)=\int_{\mathbb{R}}(x-\bar x(0))^2p(x,0)dx<+\infty,
\end{equation}
where $p(x,0)=\nu(x,0)/N(0)$, and we consider the PDE
\begin{equation}\label{PDE_SM}
\dot\nu(x,t)=m(\nu,x)\nu(x,t)+\frac{\mu}{2}\Delta\nu(x,t).
\end{equation}

@@

Here we show that solutions to the PDE described in the main text are twice continuously differentiable with respect to space and differentiable with respect to time. Our approach is an adaptation of that taken by Volpert [-@Volpert2014].

Recall that the PDE we consider is of the form
\begin{equation}\label{PDE_supp}
\dot\nu(x,t)=m(\nu,x)\nu(x,t)+\frac{\mu}{2}\Delta\nu(x,t),
\end{equation}
where $m(\nu,x)$ bounded above by $r\in\mathbb{R}$ is shorthand for $m((K\nu)(x,t),x)$ and $(K\nu)(x,t)=\int_\mathbb{R}\kappa(x,y)\nu(y,t)dy$ for some non-negative, bounded $\kappa$. Denoting $L^1(\mathbb{R})$ the Lebesgue space of integrable functions on $\mathbb{R}$ (ie, the set of functions $f:\mathbb{R}\to\mathbb{R}$ that satisfy $\|f\|_1=\int_\mathbb{R}|f(x)|dx<+\infty$ under the equivalence relation $f=g$ iff $\|f-g\|_1=0$), we assume $\nu(\cdot,0)\in L^1(\mathbb{R})$. Under these assumptions, we have already proved in SM \S\ref{finite} that $\nu(\cdot,t)\in L^1(\mathbb{R})$ for all $t\geq0$. Before proving the regularity of $\nu(x,t)$, we review the approach taken by Volpert [-@Volpert2014].

@@

In chapter 9 \S2.2 of Volpert [-@Volpert2014] complimentary conditions are assumed where $\kappa(x,y)=\varphi(x-y)$ for some integrable function $\varphi$ and initial condition $\nu(\cdot,0)\in L^\infty(\mathbb{R})$, where $L^\infty(\mathbb{R})$ is the Lebesgue space of bounded functions on $\mathbb{R}$ (ie, the set of functions $f:\mathbb{R}\to\mathbb{R}$ satisfying $\|f\|_\infty=\sup_{x\in\mathbb{R}}|f(x)|<+\infty$ under the equivalence relation $f=g$ iff $\|f-g\|_\infty=0$). Instead of the growth term $m(\nu,x)\nu(x,t)$ of PDE (\ref{PDE_supp}), Volpert worked with $f(\nu(x,t),(K\nu)(x,t))$ for some function $f:\mathbb{R}^2\to\mathbb{R}$ differentiable with respect to both arguments and satisfying, for some $q\geq0$,
\begin{equation}
f(x,y)\leq q(1+|x|+|y|), \ \forall \ x,y\in\mathbb{R}.
\end{equation}
Under these conditions, Volpert used semigroup theory to prove the solution $\nu(x,t)$ takes values in $L^\infty(\mathbb{R})$ for all $t\geq0$ and is differentiable with respect to time and twice continuously differentiable with respect to space. In particular, defining $\nu(t)$ as the function $x\to\nu(x,t)$ and keeping $\Phi(x,t)$ as defined in SM \S\ref{finite}, Volpert states the semigroup $\{S(t)\}_{t\geq0}$ given by 
\begin{equation}
S(t)\nu(0)=\int_0^t\frac{\mu}{2}\Delta\nu(x,s)ds=\int_\mathbb{R}\Phi(x-y,t)\nu(y,0)dy, \ \ \forall \ \ t\geq0, \ \nu(0)\in L^\infty(\mathbb{R})
\end{equation}
satisfies $\|S(t)\nu(0)\|_\infty\leq\|\nu(0)\|_\infty$ for all $t\geq0$ and $\nu(0)\in L^\infty(\mathbb{R})$. By setting
\begin{equation}
F(v)=f\left(v(x),\int_\mathbb{R}\varphi(x-y)v(y)dy\right), \ \ \forall \ \ v\in L^\infty(\mathbb{R}),
\end{equation}
Volpert notes that $F$ is locally Lipschitz continuous on $L^\infty(\mathbb{R})$ and in particular $\|F(v)\|_\infty\leq q\|v\|_\infty$. Hence, under Volperts conditions, solutions to PDE (\ref{PDE_supp}), given they exist, can be written as
\begin{equation}\label{volp_sol}
\nu(t)=S(t)\nu(0)+\int_0^tS(t-s)F(\nu(s))ds \ \ \forall \ \ t\geq0, \ \nu(0)\in L^\infty(\mathbb{R}).
\end{equation}
Volpert interprets the right-hand-side of equation (\ref{volp_sol}) as an operator $T$ acting on functions $v\in L^\infty(\mathbb{R}\times[0,\tau])$ with $\tau>0$. The space $L^\infty(\mathbb{R}\times[0,\tau])$ is given the norm $\|v\|_\infty^\infty=\max_{t\in[0,\tau]}\|v(t)\|_\infty$. The operator $T$ is claimed to have a single fixed point for sufficiently small $\tau$, which means that there exists exactly one function $v$ such that $Tv=v$. That is, there exists $v\in L^\infty(\mathbb{R}\times[0,\tau])$ such that
\begin{equation}
v(t)=(Tv)(t)=S(t)v(0)+\int_0^tS(t-s)(Fv)(s)ds\in L^\infty(\mathbb{R}) \ \ \forall \ \ t\in[0,\tau].
\end{equation}
Volpert merely states this is easy to conclude and the regularity conditions follow from it. We expand on why this is so.

Combining the results $\|S(t)v(s)\|_\infty\leq\|v(s)\|_\infty$, $\|Fv(s)\|_\infty\leq q\|v(s)\|_\infty$ for all $t\geq0$, $s\in[0,\tau]$ and $v\in L^\infty(\mathbb{R}\times[0,\tau])$ with Minkowski's inequality, we have
\begin{equation}
\|Tv\|_\infty^\infty\leq\|v\|_\infty^\infty+q\tau\|v\|_\infty^\infty=(1+q\tau)\|v\|_\infty^\infty.
\end{equation}
Hence, for each $\tau>0$, $T$ is a contraction mapping on $L^\infty(\mathbb{R}\times[0,\tau])$. Banach's fixed point theorem therefore implies $T$ has exactly one fixed point. This holds for all $\tau>0$ which implies the existence and uniqueness of solutions to PDE (\ref{PDE_supp}) under Volpert's conditions. Hence, the solutions are differentiable with respect to time. Since $S(t)v$ is twice differentiable with respect to space for all $t\geq0$ and since $f$ is differentiable with respect to both arguments, $\int_0^tS(t-s)(Fv)(s)ds$ is twice differentiable with respect to space. Hence, solutions to PDE (\ref{PDE_supp}) are twice differentiable with respect to space.

@@

To adapt Volperts approach to our situation, note that $\|S(t)\nu(0)\|_1=\|\nu(0)\|_1$ for all $\nu(0)\in L^1(\mathbb{R})$ and, setting
\begin{equation}
(Fv)(x)=m\left(\int_\mathbb{R}\kappa(x,y)v(y)dy,x\right)v(x), \ \ \forall \ \ v\in L^1(\mathbb{R}),
\end{equation}
we have $\|Fv\|_1\leq r\|v\|_1$ for all $v\in L^1(\mathbb{R})$. Hence, when they exist, solutions to PDE (\ref{PDE_supp}) under our assumptions can be written as
\begin{equation}
\nu(t)=S(t)\nu(0)+\int_0^tS(t-s)(F\nu(s))ds \ \ \forall \ \ t\geq0, \ \nu(0)\in L^1(\mathbb{R}).
\end{equation}
Then, using the same arguments employed by Volpert [-@Volpert2014], we conclude solutions to PDE (\ref{PDE_supp}) under our conditions exist, are unique and exhibit the desired regularity conditions.
